---
title: "Answering Business Questions using SQL"
author: "Yassir"
date: "2023-04-01"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

- import the RSQLite and DBI libraries.
```{r}
library(DBI)
library(RSQLite)
library(tidyverse)
```


- Create a run_query() function that takes in a SQL query as an argument and returns the result in a dataframe

- Create a show_tables() function that calls the run_query() function to return a list of all tables and views in the database
```{r}
db <- 'chinook.db'
run_query <- function(q) {
  conn <- dbConnect(SQLite(), db)
  result <- dbGetQuery(conn, q)
  dbDisconnect(conn)
  return(result)
}
show_tables <- function() {
  q = "SELECT name, type FROM sqlite_master WHERE type IN ('table', 'view')"
  return(run_query(q))
}
show_tables()
```
1. Write a query that returns each genre, with the number of tracks sold in the USA:
-   in absolute numbers
-   in percentages
```{r}
purchasing_albums = '
WITH tracks_sold_usa AS
   (
    SELECT il.* FROM invoice_line il
    INNER JOIN invoice i on il.invoice_id = i.invoice_id
    INNER JOIN customer c on i.customer_id = c.customer_id
    WHERE c.country = "USA"
   )
SELECT
    g.name genre,
    count(tsu.invoice_line_id) tracks_sold,
    cast(count(tsu.invoice_line_id) AS FLOAT) / (
        SELECT COUNT(*) from tracks_sold_usa
    ) percentage_sold
FROM tracks_sold_usa tsu
INNER JOIN track t on t.track_id = tsu.track_id
INNER JOIN genre g on g.genre_id = t.genre_id
GROUP BY 1
ORDER BY 2 DESC
LIMIT 10;
'
run_query(purchasing_albums)
```
2. Create a plot to show this data
```{r}
library(ggplot2)

genre_sales_usa = run_query(purchasing_albums)

ggplot(data = genre_sales_usa, 
       aes(x = reorder(genre,-percentage_sold),
           y = percentage_sold)) +
         geom_bar(stat = "identity") +
         labs(title = "genre sold in the USA",
              x = "genre",
              y = "percentage")


```

3. Interprets the data and makes a recommendation for the three artists whose albums we should purchase for the store

- The data show the percentage of sold genre in the USA
- The show that the top three rated genres are Rock, Alternative & Punk, and Metal.
- While the Chinook record store has four albums (Hip-Hop, Punk, Pop, Blues) to choose from, I recommend choosing the following three artists' albums:
  - Red Tone (Punk)
  - Slim Jim Bites (Blues)
  - Meteor and the Girls (Pop)
- Rock music represents the largest portion of sales. If we aim to obtain a more comprehensive understanding of sales in the USA, it could be beneficial to inquire with the record label about any emerging rock bands. 

## Analyzing Employee Sales Performance

1. Write a query that finds the total dollar amount of sales assigned to each sales support agent within the company. Add any extra attributes for that employee that you find are relevant to the analysis.

```{r}
employee_performance = '
WITH customer_invoice AS
(
 SELECT i.customer_id,
        c.support_rep_id,
        SUM(i.total) as total
   FROM invoice as i
  INNER JOIN customer c ON i.customer_id = c.customer_id
  GROUP BY 1,2
)
SELECT
    e.first_name || " " || e.last_name as employee,
    e.hire_date,
    SUM(ci.total) as total_sales
FROM customer_invoice as ci 
INNER JOIN employee e ON e.employee_id = ci.support_rep_id
GROUP BY 1;
'
run_query(employee_performance)

```
2. Create a plot of the results of your query
```{r}
library(ggplot2)
employee_sales = run_query(employee_performance)
ggplot(data = employee_sales,aes(x = reorder(employee, -total_sales), y = total_sales)) +
  geom_bar(stat = "identity") +
  labs(title = "Employee Sales Performance",
              x = "employee",
              y = "total sales")
```


3. Write a short statement describing your results, and provide a possible interpretation.
- The data show employee sales performance
- Jane Peackock is the top performer  of sold genre in the USA
- Jane's high performance could be partly due to her extended tenure at the company. 
- To truly gauge employee productivity, it may be beneficial to standardize sales based on the number of days or hours worked.

## Analyzing Sales by Country

1. Write a query that collates data on purchases from different countries. For each country, include:
  - total number of customers
  - total value of sales
  - average value of sales per customer
  - average order value
2. Where a country has only one customer, collect them into an "Other" group.

3. The results should be sorted by the total sales from highest to lowest, with the "Other" group at the very bottom. 

```{r}
country_wise_sales =
'with customer_invoice AS
(
SELECT CASE
           WHEN (
                 SELECT count(*)
                 FROM customer
                 where country = c.country
                ) = 1 THEN "Other"
           ELSE c.country
       END AS country,
       c.country as country,
       count(DISTINCT c.customer_id) as total_customers,
	   SUM(il.unit_price) as total_sales,
	   SUM(il.unit_price) / count(distinct i.customer_id) as average_sales_per_customer,
	   SUM(il.unit_price) / count(DISTINCT il.invoice_id) as average_order_value
  FROM customer as c
  JOIN invoice as i
    ON c.customer_id = i.customer_id
  JOIN invoice_line as il
    ON il.invoice_id = i.invoice_id
  GROUP BY 1
  ORDER BY 2 DESC
)  
  
SELECT country,
       total_customers,
	   total_sales,
	   average_sales_per_customer,
	   average_order_value,
	   CASE
            WHEN country = "Other" THEN 1
            ELSE 0
        END AS sort
	   FROM customer_invoice
	   GROUP by 1
	   ORDER BY sort ASC, total_sales DESC'
run_query(country_wise_sales)
	   
```
## Visualizing Sales by Country

1. For each dimension, create a visualization which demonstrates the data we collated in the previous step.

  - You should decide whether the "Other" group is relevant to your analysis and make decisions on where to include it (if anywhere) in your visualizations.

```{r}
library(ggplot2)
sales_by_country = run_query(country_wise_sales)

ggplot(data = sales_by_country,aes(x = reorder(country, -total_customers), y = total_customers, fill = country)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Customers by Countries",
              x = "countries",
              y = "total customers") +
  theme(panel.background = element_rect(fill = "lightgrey")
) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(data = sales_by_country, aes(x = reorder(country, -total_sales), y = total_sales, 
fill = country)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Total Sales by Country",
    x = "Country",
    y = "Total Sales"
  ) + 
theme(panel.background = element_rect(fill = "lightgrey")
) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(data = sales_by_country, aes(x = reorder(country, -average_sales_per_customer), y = average_sales_per_customer, 
fill = country)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Average Sales per customer by Country",
    x = "Country",
    y = "Average Sales per Customer"
  ) + 
theme(panel.background = element_rect(fill = "lightgrey")
) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(data = sales_by_country, aes(x = reorder(country, -average_order_value), y = average_order_value, 
fill = country)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Average Order Value by Country",
    x = "Country",
    y = "Average Order Value"
  ) + 
theme(panel.background = element_rect(fill = "lightgrey")
) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
3. Write a few sentences interpreting your data and visualizations, and make one or more recommendations to the marketing team on which countries have potential for growth.

- Based on the data visualization, our three most important markets,excluding others, are USA, Canada, and Brazil. However, Czech Republic, Portugal, and India are promising markets in terms of average sales per customers. Again, Czech Republic, The UK, and India are valuable markets with respect to average order value. Thus, the marketing team should consider these markets when developing their marketing plans.


## Albums vs. Individual Tracks

1. Write a query that categorizes each invoice as either an album purchase or not, and calculates the following summary statistics:
  - Number of invoices
  - Percentage of invoices
  
```{r}
invoice_vs_track = '
WITH invoice_track AS
(SELECT invoice_id,
MAX(track_id) AS track_id
FROM invoice_line
GROUP BY invoice_id
)
SELECT
album_purchase,
COUNT(invoice_id) AS invoices,
ROUND(CAST(COUNT(invoice_id) AS FLOAT) / (SELECT COUNT(*)
FROM invoice), 2) AS percentage
FROM (SELECT it.*,
CASE WHEN (SELECT t.track_id
FROM track AS t
WHERE album_id = (
SELECT t2.album_id FROM track AS t2
WHERE t2.track_id = it.track_id
)
EXCEPT
SELECT il2.track_id
FROM invoice_line AS il2
WHERE il2.invoice_id = it.invoice_id
) IS NULL
AND
(SELECT il2.track_id
FROM invoice_line AS il2
WHERE il2.invoice_id = it.invoice_id
EXCEPT
SELECT t.track_id
FROM track AS t
WHERE t.album_id = (
SELECT t2.album_id FROM track AS t2
WHERE t2.track_id = it.track_id
)
) IS NULL
THEN "Yes"
ELSE "NO"
END AS album_purchase
FROM invoice_track AS it
)
GROUP BY album_purchase;'

run_query(invoice_vs_track)
```
2. Write one to two sentences explaining your findings, and making a prospective recommendation on whether the Chinook store should continue to buy full albums from record companies

  - Based on the findings, album purchases make up 19% of the overall sales.Thus, it is not recommended to change switch the strategy. Rather, management should keep adopting the current strategy of purchasing all tracks in albums, given its great contribution to increasing revenues.






