---
title: "Bike Sharing Project: Applying Multiple Linear Regression"
author: "Yassir"
date: "2024-03-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# loading tidyverse
library(tidyverse)
```

```{r}
# loading the dataset
bike_df <- read_csv("day.csv")

```
```{r}
# exploring the dataset
glimpse(bike_df)

```

```{r}
# checking the summary statistics
summary(bike_df)
```

```{r}
# checking missing values
colSums(is.na(bike_df))
```
```{r}
# selecting numeric variables for correlations 
bike_df_numeric <- select_if(bike_df, is.numeric)
head(bike_df_numeric)
```

```{r}
# let's remove factors that are not related
bike_df_numeric <- select(bike_df_numeric,
                          - casual, - instant, - registered)
head(bike_df_numeric)
```


```{r}
# let's do correlations before performing regressions
cor(bike_df_numeric)
```

# the correlation matrix clarifies that cnt is moderlately corelated with the following predictors: season , year, weathersit, temp, and atem

```{r}
# selecting our variables of interest and applying cor again
bike_var_df <- bike_df_numeric %>%
  select(season, yr, weathersit, temp, atemp, cnt)

head(bike_var_df)
```

```{r}
model_1 <- lm(cnt ~ season + yr + weathersit + temp + atemp, data = bike_var_df)

summary(model_1)
```

```{r}
# loading "car" package to check vif
library(car)
```

```{r}
# checking vif
vif_values <- vif(model_1)
print(vif_values)
```

```{r}
# let's remove atemp from the model
model_2 <- lm(cnt ~ season + yr + weathersit + temp, data = bike_var_df)

summary(model_2)
```

```{r}
# checking vif
vif_values <- vif(model_2)
print(vif_values)

```
# While "temp" has very little multicollinearity with the other independent variables, model_2 seems to be good with Adjusted R-squared almost 0.78.

```{r}
# visualizing the residuals to evaluate the model

residuals_df <- data.frame(residuals = model_2$residuals)

# ploting the residuals
ggplot(data = residuals_df,
aes(x = residuals)) +
geom_histogram()
```
The residuals follow a normal distribution. Hence our assumption for Linear Regression is valid.
