---
title: 'Guided Project: Investigating Fandango Movie Ratings'
author: "Yassir"
date: "2023-09-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Is Fandango continuing to artificially boost ratings?
In October 2015, Walt Hickey of FiveThirtyEight released a well-received article, accessible [link](https://fivethirtyeight.com/features/fandango-movies-ratings/), in which he provided compelling proof indicating potential bias and lack of transparency within Fandango's movie rating system. In this undertaking, we will examine more current movie rating data to ascertain if there have been any alterations to Fandango's rating system since Hickey's investigation.

# Understanding the Data
```{r}
# loading tidyverse
library(tidyverse)

# reading the two data sets
fandango_score_comparison <- read_csv("fandango_score_comparison.csv")

movie_ratings_16_17 <- read_csv("movie_ratings_16_17.csv")

# exploring the data sets

dim(fandango_score_comparison) # 146 rows with 22 columns
dim(movie_ratings_16_17) # 214 rows with 15 columns

# selecting relevant columns from fandango_score_comparison data set
fandango_previous <- fandango_score_comparison %>%
  select(FILM, Fandango_Stars, Fandango_Ratingvalue, Fandango_votes, Fandango_Difference)

head(fandango_previous)

# selecting movie, year, and fandango columns for movie_ratings_16_17 data set
fandango_after <- movie_ratings_16_17 %>%
  select(movie, year, fandango)
head(fandango_after)

```

Our objective is to assess whether Fandango's rating system underwent any changes subsequent to Hickey's analysis. The target population for our investigation encompasses all the movie ratings stored on Fandango's website, irrespective of their release year.

In order to examine whether the characteristics of this population evolved following Hickey's analysis, we intend to sample the population at two distinct time points: one prior to the analysis and one subsequent to it, enabling us to make a meaningful comparison.

The data we are utilizing was sampled precisely when we needed it: one sample was obtained prior to the analysis, and the other was taken after. It is imperative that these samples accurately represent the population; otherwise, we risk encountering significant sampling errors that could lead to incorrect conclusions.

Hickey's article and the README.md of the [dataset repository](https://github.com/fivethirtyeight/data/tree/master/fandango) reveal that he employed specific criteria for sampling:

- Movies must have had a minimum of 30 fan ratings on Fandango's website at the time of sampling (Aug. 24, 2015).
- Movies must have had tickets on sale in 2015.

This sampling process was evidently not random, as not every movie had an equal chance of being included in the sample. Some movies had no chance at all, such as those with fewer than 30 fan ratings or those without tickets on sale in 2015. Consequently, there are doubts about whether this sample accurately represents the entire population we seek to describe. It appears more likely that it does not, primarily due to the influence of temporal trends. For instance, movies in 2015 might have been exceptionally good or bad compared to other years.

The sampling criteria for our other sample can be found in the README.md of the [dataset's repository](https://github.com/mircealex/Movie_ratings_2016_17):

- Movies must have been released in 2016 or later.
- Movies must have garnered a substantial number of votes and reviews, although the exact number remains unclear from the README.md or the data.

Similar to the first sample, this second sample is also susceptible to temporal trends and is unlikely to be representative of our population of interest.

Both authors had specific research inquiries in mind when conducting their sampling, and they employed a set of criteria to obtain samples that aligned with their research questions. This approach is known as purposive sampling, and while it served their research purposes well, it may not be particularly useful for our own analysis.

# Changing the Goal of our Analysis
At this point, we can either collect new data or change our the goal of our analysis. We choose the latter and place some limitations on our initial goal.

Instead of trying to determine whether there has been any change in Fandango's rating system after Hickey's analysis, our new goal is to determine whether there's any difference between Fandango's ratings for popular movies in 2015 and Fandango's ratings for popular movies in 2016. This new goal should also be a fairly good proxy for our initial goal.

# Isolating the Samples We Need
With this new research goal, we have two populations of interest:

1. All Fandango's ratings for popular movies released in 2015.
1. All Fandango's ratings for popular movies released in 2016.

We need to be clear about what counts as popular movies. We'll use Hickey's benchmark of 30 fan ratings and count a movie as popular only if it has 30 fan ratings or more on Fandango's website.

Although one of the sampling criteria in our second sample is movie popularity, the `fandango_after` dataframe doesn't provide information about the number of fan ratings. We should be skeptical once more and ask whether this sample is truly representative and contains popular movies (movies with over 30 fan ratings).

One quick way to check the representativity of this sample might be to sample randomly 10 movies from it and then check the number of fan ratings ourselves on Fandango's website. 

```{r}
set.seed(1)
sample_n(fandango_after, size = 10)
```
In the code above, I employed a random seed value of 1. This practice is considered sound because it indicates that I did not experiment with various random seeds solely to obtain a favorable sample.

while inspecting the fan ratings for the aforementioned movies, the website is not available outside the USA. However, I found out through the solution guide that as of August 2019, Fandango has transitioned away from utilizing the 5-Star Fan Ratings previously mentioned. Instead, Fandango has adopted the Rotten Tomatoes verified Audience Score. 

# Let's isolate the movies released in 2015 and 2016
```{r}
# let's start by extracting the year 2015 from the film column strings
fandango_previous <- fandango_previous %>% 
  mutate(year = str_sub(FILM, -5, -2))

# let's see the frequency distribution
table(fandango_previous$year)

# another alternative way to see the freq. distribution
fandango_previous %>%
  group_by(year) %>%
  summarize(freq = n())

# let's isolate the 2015 data set
fandango_2015 <- fandango_previous %>% 
  filter(year == 2015)
table(fandango_2015$year)

# let's now isolate the movies in other data set
# first we see the frequency distribution
table(fandango_after$year)

# let's now isolate the fandango_2016 dataset
fandango_2016 <- fandango_after %>%
  filter(year == 2016)
table(fandango_2016$year)

# We have now two isolated data sets: fandango_2015 & fandango_2016
```

# Comparing Distribution Shapes for 2015 and 2016
```{r}
# To find if there is any difference between the two samples, let's see the shape of the two distributions using kernel density plots.

ggplot(data = fandango_2015, aes(x = Fandango_Stars)) +
geom_density() +
  geom_density(data = fandango_2016,
               aes(x = fandango, color = "blue")) +
  labs(title = "Comparing distribution shapes for Fandango's ratings\n(2015 vs 2016)",
       x = "Stars",
       y = "Density") +
  scale_x_continuous(breaks = seq(0, 5, by = 0.5), 
                     limits = c(0, 5))

```

The figure above reveals two notable observations:

Both distributions exhibit a pronounced leftward skew.
In comparison to the 2015 distribution, the 2016 distribution is marginally shifted to the left.
The leftward skewness implies that the majority of movies on Fandango receive predominantly high and very high fan ratings. Considering that Fandango is also a ticket-selling platform, these high ratings raise some skepticism. A thorough investigation of this phenomenon would be intriguing.

The slight leftward shift in the 2016 distribution holds particular significance for our analysis. It indicates that movie ratings in 2016 were slightly lower in comparison to those in 2015. This suggests a discernible difference between Fandango's ratings for popular movies in 2015 and those in 2016. Furthermore, it clarifies the direction of this difference: the ratings in 2016 were marginally lower when contrasted with those in 2015.

# Comparing Relative Frequencies
```{r}
# let's use relative frequencies instead of absolute frequencies:
  # fandango_2015

fandango_2015 %>% 
  group_by(Fandango_Stars) %>% 
  summarize(Freq = n()) %>%
  mutate(Percentage = Freq / nrow(fandango_2015) * 100)

  # fandango_2016
fandango_2016 %>%
  group_by(fandango) %>%
  summarize(Freq = n()) %>%
  mutate(Percentage = Freq / nrow(fandango_2016) *100)

```

From the above the relative frequencies, In 2016, we observed lower percentages of very high ratings (4.5 and 5 stars) when compared to 2015. Specifically, in 2016, less than 1% of the movies achieved a perfect 5-star rating, whereas in 2015, this percentage was nearly 7%. Additionally, the prevalence of 4.5-star ratings was higher in 2015, with approximately 13% more movies receiving this rating compared to 2016.

Furthermore, the minimum rating in 2016 was lower, set at 2.5 stars, in contrast to the 3-star minimum of 2015. This clear discrepancy between the two frequency distributions is evident.

Conversely, for some other rating categories, we observed an increase in the percentage of movies in 2016. There was a notable rise in the proportion of films receiving 3.5 and 4 stars compared to 2015. Both 3.5 and 4.0 stars represent high ratings, which introduces a challenge to the observed direction of change as depicted in the kernel density plots.

# Determining the Direction of the Change
```{r}
# Let's take some summary metrics to understand the direction of the change. I will compute the mean, the median, and the mode for both distributions and then use a bar graph to plot the values.

# let's use the created function of the mode 
mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

summary_2015 <- fandango_2015 %>% 
  summarize(year = "2015",
            mode = mode(Fandango_Stars),
            mean = mean(Fandango_Stars),
            median = median(Fandango_Stars))
print(summary_2015)

summary_2016 <- fandango_2016 %>% 
  summarize(year = "2016",
            mode = mode(fandango),
            mean = mean(fandango),
            median = median(fandango))
print(summary_2016)

# let's combine these two 2015 & 2016 summary dataframes
summary_df <- bind_rows(summary_2015, summary_2016)
print(summary_df)

# let's build the grouped bar chart
# let's first reshape the data using gather function 
summary_df <- summary_df %>%
  gather(key = "statistic", value = "value", - year)
print(summary_df)


ggplot(data = summary_df, aes(
  x = statistic, y = value, fill = year
)) +
  geom_bar(stat = "identity", position = "dodge") + labs(title = "Comparing summary statistics: 2015 vs 2016",
       x = "",
       y = "Stars")
```

# Conclusion
Our examination has confirmed a subtle contrast between Fandango's assessments of popular movies in 2015 and those in 2016. Additionally, we found that, on average, well-received films from 2016 received lower ratings on Fandango compared to those from 2015.

While the exact reason for this shift cannot be definitively determined, it is highly likely that it resulted from Fandango addressing the bias in its rating system following Hickey's analysis.