---
title: 'Guided Project: New York Solar Resource Data Using APIs'
author: "Yassir"
date: "2023-07-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

## The goal of the project is to extract the solar resource data using APIs for New York City in JSON format and convert it into a dataframe. 

## The final dataframe will include three variables,namely, month, avg_dni, avg_ghi, and avg_lat_tilt

# Finding the Suitable Endpoint and Parameters to Query the API
```{r}
# Store your API key in a variable
api_key <- "LPbNz57HixXexZBSvxUZvIB4UXOKe2od7BMRUsUr"

# Identify and store the API URL in Json format
url <- "https://developer.nrel.gov/api/solar/solar_resource/v1.json?api_key=DEMO_KEY&lat=40&lon=-105"

# Store the three parameters to request the New York City solar data as a list

parameters_list <- list(key = api_key, lat = 41, lon = -75)
```

# Extracting the New York Solar Resource Data
```{r}
# Loading the httr package
library(httr)

# Use the GET() function to request the data from the API with url and parameters_list

response <- GET(url, query = parameters_list)

# Track errors
status_code(response)
http_status(response)
http_type(response)

# Extract the API response content as text
content <- content(response, "text")

# Display this content, and check it
print(content)
```
# Parsing the JSON into an R Object

```{r}
# Parse the API response content to an R object using the jsonlite::fromJSON() function
# Store the result as json_lists

json_lists <- jsonlite::fromJSON(content)

# Display the structure of the R object using the str() function
str(json_lists)
```
# How to Create a Dataframe from a Complex List
# Building a Dataframe from a Complex List
```{r}
# Extract the outputs list from json_lists
outputs_list <- json_lists$outputs

# Extract the monthly vector (monthly) from the following lists: avg_dni, avg_ghi, avg_lat_tilt
avg_dni <- outputs_list$avg_dni$monthly
avg_ghi <- outputs_list$avg_ghi$monthly
avg_lat_tilt <- outputs_list$avg_lat_tilt$monthly

# Combine the monthly vectors into a dataframe using the tibble::tibble() function
## Add the month column containing month abbreviations: Jan, Feb, . . . ,Dec
### The month column should appear on the left side of the dataframe
## Save the result as dataframe
dataframe <- tibble::tibble("month" = month.abb, 
                            "avg_dni" = avg_dni, 
                            "avg_ghi" = avg_ghi,
                            "avg_lat_tilt" = avg_lat_tilt)

# Display the dataframe
print(dataframe)

# What do you notice, and what impact can this have on future uses of this dataframe?
# The variables of the dataframe are still lists and they need to be converted into numeric data

# To convert a dataframe that contains variables stored as lists into a regular dataframe structure, we can use the tidyr::unnest() function from the tidyr package

library(tidyverse)
df <- dataframe
print(df)
df_unnested <- unnest(df)
print(df_unnested)
# let's follow the instruction of the guided project as they provide another solution for reaching the same dataframe above using the unnest function
```
# Extracting a Dataframe from a Complex List
```{r}
# Extracting the outputs list
outputs_list <- json_lists$outputs

# Simplifying the outputs list
simplified_outputs_list <- unlist(outputs_list)

# Restructuring the simplified list into a matrix of 13 rows (the annual value and 12 months values)
data_matrix <- matrix(data = simplified_outputs_list, nrow = 13)

# Removing the annual values from the data matrix
data_matrix <- data_matrix[-1, ]

# Converting the matrix into a dataframe using the `as.data.frame()` function
dataframe_2 <- as.data.frame(data_matrix)

# Displaying the dataframe
print(dataframe_2)

# What difference do you see between the first approach and this one?
# It is noticed that while the dataframe is numeric, the month column is still missing.
```

#  Putting It All together
```{r}
# Load the packages: httr and dplyr
library(httr)
library(dplyr)

# Create the custom nrel_api_json_get_df() function based on what we did in previous lessons.
  ## Copy-paste and modify it accordingly for this purpose
  ## Convert the content into a dataframe
nrel_api_json_get_df <- function(endpoint, queries = list()) {
  ## Preparing the URL 
  url <- modify_url("https://developer.nrel.gov", path = endpoint)
  ## Querying the API
  response <- GET(url, query = queries)
  ## Tracking errors
  if ( http_error(response) ){
    print(status_code(response))
    print(http_status(response))
    stop("Something went wrong.", call. = FALSE)
  }
  if (http_type(response) != "application/json") {
    stop("API did not return json", call. = FALSE)
  }
  ## Extracting content
  json_text <- content(response, "text")
  ## Converting content into Dataframe
  table_lst <- jsonlite::fromJSON(json_text)
  dataframe <- tibble::tibble("month" = month.abb,
                              "avg_dni" = as.numeric(table_lst$outputs$avg_dni$monthly),
                              "avg_ghi" = as.numeric(table_lst$outputs$avg_ghi$monthly),
                              "avg_lat_tilt" = as.numeric(table_lst$outputs$avg_lat_tilt$monthly))
  ## Returning the dataframe  
  dataframe
}
# Use the custom nrel_api_json_get_df() function to extract the solar resource as a dataframe
## Providing the `"api/solar/solar_resource/v1.json"` as the `endpoint` parameter
## Providing the `parameters_list` variable as `queries` parameter
solar_resource_df <- nrel_api_json_get_df("api/solar/solar_resource/v1.json?api_key=DEMO_KEY&lat=40&lon=-105", parameters_list)
# Printing the output dataframe
solar_resource_df
print(solar_resource_df)
```

Visualizing New York City Solar Resource Data
```{r}
# Load the ggplot2 and dplyr packages.
library(ggplot2)
library(dplyr)

# Create the first plot with the following settings:
  ## Use the ggplot() function to plot (for each month) the average direct irradiation (avg_dni) from solar_resource_df
  ## Use geom_line() and geom_point() functions to get a line plot with a dot for each value
  ## Use the theme_bw() theme.
ggplot(data = solar_resource_df, aes(x = month, y = avg_dni, group = 1))+
  geom_line() +
  geom_point() +
  theme_bw()

# Create the second plot with the following settings:
  ## Convert the month column into a factor using the following command: solar_resource_df <- solar_resource_df %>% mutate(month = factor(month, levels = month.abb))

  ## Reproduce the first plot with the same settings. You can copy-paste the first plot code snippet

solar_resource_df <- solar_resource_df %>% mutate(month = factor(month, levels = month.abb))

ggplot(data = solar_resource_df, aes(x = month, y = avg_dni, group = 1))+
  geom_line() +
  geom_point() +
  theme_bw()

# What do you notice comparing both plots?
  ## What have you learned about this command solar_resource_df <- solar_resource_df %>% mutate(month = factor(month, levels = month.abb))?

# I noticed that the ordering of the x-axis in the first plot is alphabetical, whereas in the second plot, it follows a chronological order from January to December. This enables us to arrange the labels on the plot according to our preferences.


```