---
title: 'Guided Project: Creating An Efficient Data Analysis Workflow'
author: "Yassir Al Masoudi"
date: "2022-10-08"
output: html_document
---

# The project's objective is to find the most profitable book.

# Loading some libraries 
```{r}
library(tidyverse)
library(dplyr)  
library(readr) # reading the dataset
```
# Exploring and understanding how big the dataset is!
```{r}
#loading the dataset from the book_reviews file csv
book.df <- read.csv("book_reviews.csv") 

View(book.df) # Exploring the dataset

head(book.df) # displaying the first 6 rows of the dataset

tail(book.df) # displaying the last 6 rows of the dataset

dim(book.df) # displaying the dimensions of the dataset (2000 rows, 6 columns)

colnames(book.df) # displaying column names (book, review, state, price). 

typeof(book.df$book) # The book column has a character data type
typeof(book.df[["book"]]) # Another way to check the data type in a column
typeof(book.df$review) # The review column has a character data type
typeof(book.df$state) # The state column has a character data type
typeof(book.df$price) # The price column has a double data type

# using for loop to find out the columns data type
for (i in colnames(book.df)) {
  print(typeof(book.df[[i]]))
}

unique(book.df$book) # Return the book column data without duplicates. The book column has the following data: ("R Made Easy","R For Dummies","Secrets Of R For Advanced students","Top 10 Mistakes R Beginners Make", "Fundamentals of R For Beginners"). The book column has not missing values.
unique(book.df$review) # Return the review column data, indicating reviewers' scores. The review column has the following data: ("Excellent","Fair","Poor","Great", NA,"Good"). The review column has missing values.
unique(book.df$state) # Return the state column where the book was purchased. The state column has the following data: ("TX", "NY","FL", "Texas","California"). The state column has no missing values.
unique(book.df$price) # Return the price column, showing the prices of the book. The price column has the following data: (19.99 15.99 50.00 29.99 39.99). The price column has no missing values.

# Using for loop to check unique column values
for (c in colnames(book.df)) {
  print("Unique column values:")
  print(c)
  print(unique(book.df[[c]]))
  print("__________________")
}

glimpse(book.df) # showing an overview of the dataset
```
# Cleaning the dataset by finding and removing missing data
```{r}
# The review column has missing values. So, we need to remove them by using the filter, !, & is.na functions.
filtered_review <- book.df %>%
  filter(!(is.na(review)))
glimpse(filtered_review) # displaying the new dataset after removing the missing data
dim(filtered_review) # checking the new dimensions of the dataset after removing the missing data. The new dataset has 1794 rows and 4 columns
dim(book.df) # recalling the old dataset has 2000 rows and 4 columns
Removed_data_size <- 2000-1794 # 206 is the size of the removed data
Percentage_removed_data_size <- 206/2000 # .103 or 10% reviews were removed from the original data. The 10% won't affect the findings of any calculations we end up doing!
filtered_review #current dataset
```
# Cleaning the dataset by fixing the labelling of the state column
```{r}
unique(filtered_review$state) #let's check the inconsistency of the state column values
# let's impute a column "state" to have consistent labels of states.
imputation_filtered_review <- filtered_review %>%
  mutate(state = case_when(
    state == "TX" ~ "Texas",
    state == "NY" ~ "New York",
    state == "FL" ~ "Florida",
    state == "CA" ~ "California",
    TRUE ~ state)
  )
unique(imputation_filtered_review$state) # The labels of the state column values are now consistent after imputing a new state column
imputation_filtered_review # current dataset
```
# Converting the review column values "character" into scoring values
```{r}
#let's add another column "review_num" to include scoring values of integer types
scoring <- imputation_filtered_review %>%
  mutate(review_num = case_when(
    review == "Poor" ~ 1,
    review == "Fair" ~ 2,
    review == "Good" ~ 3,
    review == "Great" ~ 4,
    review == "Excellent" ~ 5)
  )  %>% select(book, -review, review_num, price)

head(scoring) # The review_num column has scoring integer values

#Creating a new column in the dataset called is_high_review, denoting whether or not the review has a high score or not 
review_scoring <- scoring %>%
  mutate(is_high_review = case_when(
    review_num == 4 ~ "High",
    review_num == 5 ~ "High",
    TRUE~ "Not High") 
  )
head(review_scoring$is_high_review)# is_high_review column indicates if score is high or not
cleaned_book.de <- (review_scoring) # Reassigning the dataset
head(cleaned_book.de) # the dataset is clean and ready for the analysis
View(cleaned_book.de)
```
# Finding the most profitable book
```{r}
#Grouping based on books and counting the number of purchases for each book
profitable_book <- cleaned_book.de %>%
  group_by(book) %>%
  summarize(purchased = n()) %>%
  arrange(-purchased)
head(profitable_book)
```
# Based on the analysis, it is clear that **Fundamentals of R For Beginners** is the most profitable book...  

