---
title: "Guided Project: Creating An Efficient Data Analysis Workflow, Part 2"
author: "Yassir Al Masoudi"
date: "2022-10-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Project background:

# A book company has provided us more data on some of its 2019 book sales, and it wants us to extract some usable knowledge from it. It launched a new program, encouraging customers to buy more books on July 1st, 2019, and it wants to know if this new program was successful at increasing **sales** and improving **review quality**. As the analyst, this will be your job to figure out for the guided project.

#let's load some libraries

```{r}
library(tidyverse)
library(dplyr)
library(purrr)
library(stringr)
library(lubridate)
```

#let's import the data
```{r}
sales.df <- read.csv("sales2019.csv")
```

#let's explore the dataset
```{r}
head(sales.df)
glimpse(sales.df)

#How big is the dataset? What are the column names and what do they represent?
dim(sales.df) # There are 5000 rows and 5 columns

colnames(sales.df) # date, user_submitted_review, title, total_purchased, and customer_type are all the columns' names. 

# Using for loop, What are the types of each of the columns? 

for (i in colnames(sales.df)){   #all columns are characters except total_purchased column which is integer
  print(typeof(sales.df[[i]]))
}

# Do any of the columns have missing data? If so, make a note of this.
sum(is.na(sales.df$date)) # no missing data in the date column
sum(is.na(sales.df$user_submitted_review)) # 456 missing data 
sum(is.na(sales.df$title)) # no missing data
sum(is.na(sales.df$total_purchased)) # 718 missing data
sum(is.na(sales.df$customer_type)) # no missing data
sum(is.na(sales.df)) # there are 1174 missing data
```

#let's remove the missing data in user_submitted_review
```{r}
sales.df <- sales.df %>%
  filter(!(is.na(user_submitted_review))
         )



#let's check if missing data were removed!
sum(is.na(sales.df$user_submitted_review)) # no missing data; they were removed 
dim(sales.df) # we are left with 4544 rows after removing the missing data

# Given total_purchased variable is our interest variable, let's deal with 718 missing data by finding first the aveage value of total_purchased, and then by filling in the average value in place of NA. 
total_purchased_mean <- sales.df %>%                
  filter(!is.na(total_purchased)) %>% 
  pull(total_purchased) %>% 
  mean                   

total_purchased_mean # The average of the total_purchased is around 4 

# Now, let's create a new column "total_sale" & add the **Total_puchased_mean** instead of NA
sales.df <- sales.df %>%
  mutate(
    total_sales = ifelse(is.na(total_purchased),total_purchased_mean, total_purchased) #ifelse is better suited here than if_else
  )
  
sum(is.na(sales.df$total_sales)) # no missing values
head(sales.df)
```
```{r}
#Examine the unique sentences that are present in user_submitted_review.
unique(sales.df$user_submitted_review)

#look at each of these reviews and see if you can detect specific words or phrases that help indicate if the review is positive or not.

sum(str_detect(sales.df$user_submitted_review, "Awesome"))
sum(str_detect(sales.df$user_submitted_review, "OK"))
sum(str_detect(sales.df$user_submitted_review, "Hated"))
sum(str_detect(sales.df$user_submitted_review, "not recommend"))

# Create a function to indicate if the review is positive or not
iS_positive_review <- function(review){
  review_postive <- case_when(
    str_detect(review, "Awesome") ~ TRUE,
    str_detect(review, "OK") ~ TRUE,    #OK is positive
    str_detect(review, "okay") ~ TRUE, # okay is positive
    str_detect(review, "learn") ~ TRUE,
    str_detect(review, "Never") ~ TRUE,
    TRUE ~ FALSE
    )
  return(review_postive)
}
iS_positive_review(" Never did I read a better book") # function works fo positive review

iS_positive_review(" I find the book boring") # function works for negative review

# Create a new column to indicate whether or not the review in a given row is positive or not.
sales.df <- sales.df %>%
  mutate(
    positive_review = unlist(map(user_submitted_review, iS_positive_review))
  )

sales.df %>%
  select(user_submitted_review,positive_review) # all is good! 
```
```{r}
# Convert the date column into a date and time.
# Create a new grouping column using the mutate() function to distinguish between sales that happen before July 1, 2019 and sales that happen after this date.
sales.df <- sales.df %>%
  mutate(
    sales_pre_after_july = if_else(mdy(date) < dmy("01-07-2019"), "before", "after")
  )

# Create a summary table that compares the number of books purchased before July 1, 2019 to after

sales_pre_after_program <- sales.df %>%
  group_by(sales_pre_after_july) %>%
  summarize(book_purchased = sum(total_sales)
            )
head(sales_pre_after_program)
# The number of books purchased before July 1, 2019 is 9114 compared to 9072 books purchased after July, 1 2019. Obviously, the program is not effective!
```

```{r}

# Add in the customer_type column to further subdivide the groups.
customer_groups <- sales.df %>%
  group_by(customer_type, sales_pre_after_july) %>%
  summarize(book_purchased = sum(total_sales)
            )

head(customer_groups) 
```

# Examine the results of the analysis and write about your observations?

# Based on the result, there is 1.3 % increase in sales. More specifically, the business customer group purchased around 6309 books after first of July 2019, compared to 6222 books before the first of July 2019. On the other hand, sales from individual customer group decreased from around 2892 books before July 2022 to around 2763 books, a -4% decrease. Overall, while the program was not effective in increasing the sales of books, it seems that the program slightly affected the business customer group.

```{r}
# Compare the number of positive reviews before and after July 1, 2019.


positive_review_pre_post <- sales.df %>%
  group_by(sales_pre_after_july) %>%
  summarize(count_positive_reviews = sum(positive_review)
  )

positive_review_pre_post # check the result
```
# Does it seem that review sentiment improved after the program was created? Or did it get worse?

The number of positive reviews before the program was *1134* and *1128* respectively. While the number of positive reviews slightly dropped by *6* reviews, this drop is minimal. Overall, the program was not as successful as expected. The marketing team is recommended to redesign a marketing plan and launch another program. They should further focus on those high selling books and better target their different customer groups! 
