---
title: 'Guided Project: Analyzing Forest Fire Data'
author: "Yassir Al Masoudi"
date: "2022-10-16"
output: html_document
---
## In this guided project, our task is to apply these techniques, as well as data analysis skills you gained in earlier courses, to explore a select dataset on forest fires.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
# loading some important libraries

```{r}
library(tidyverse)
```
# importing the dataset
```{r}
forest.df <- read.csv("forestfires.csv")
```
## explore the dataset ?
```{r}
dim(forest.df) # 517 observations and 13 columns

colnames(forest.df) # check columns' names
glimpse(forest.df)   # displaying an overview of the dataset
head(forest.df)     # checking first 6 raws of the dataset
sum(is.na(forest.df)) # no missing data
```

```{r}
# convert month variable into a categorical variable and make sure the months in the data are ordered correctly?

forest.df$month <- factor(forest.df$month, levels = c(
  "jan", "feb", "mar", "apr", "may", "jun", "jul", "aug",
  "sep", "oct", "nov", "dec")
)
str(forest.df) # month variable converted with 12 levels

forest.df$day <- factor(forest.df$day, levels = c(
  "mon", "tue", "wed", "thu", "fri", "sat", "sun")
)
str(forest.df) # day variable converted with 7 levels

```

```{r}
## Which months do forest fires happen the most?

# let's count the number of fires on a monthly basis
total_monthly_fires <- forest.df %>%  
  group_by(month) %>%
  summarize( total_fires = n())

# let's create a column graph 
total_monthly_fires %>% 
  ggplot(aes(x = month, y = total_fires)) +
  geom_col(fill = "#e25822") +
  theme_bw() +
  labs(
    title = "Total Monthly Fires Throughout a Year",
    x = "Months",
    y = "Number of Total Fires") +
  theme(
    plot.title = element_text(size=22),axis.text = element_text(
      size=15), axis.title=element_text(size=18))

# Based on the visualization, forest fires occur in August and September. This makes sense as forest fires occur in the summer when temperature rises!

## Which days of the week do forest fires happen the most?

total_daily_fires <- forest.df %>%  
  group_by(day) %>%
  summarize( total_fires_day = n())

# let's create a bar graph 
total_daily_fires %>% 
  ggplot(aes(x = day, y = total_fires_day)) +
  geom_col(fill = "#e25822") +
  theme_bw() +
  labs(
    title = "Total daily Fires Throughout a Week",
    x = "Days",
    y = "Number of Total Fires") + 
  theme(
    plot.title = element_text(size=22),axis.text = element_text(
      size=15), axis.title = element_text(size=18))

# Based on the visualization, forest fires mostly occur on Sundays, Saturdays, Fridays, & Mondays.
```

# look at the relationship between month and these variables
# (FFMC, DMC, DC, ISI, temp, RH, wind, and rain)

```{r}
# let's use the pivot_longer function to create a category of such variables, and then to plot multiple plots at once, using facet_wrap

forest.df_long <- forest.df  %>% 
  pivot_longer(
    cols = c(
      FFMC, DMC, DC, ISI, temp, RH, wind, rain),
    names_to = "category",
    values_to = "value"
    )
# let's visualize it, and focus on the months of aug & Sep 
forest.df_long %>%
  ggplot(aes(x = month, y = value)) +
  geom_point() +
  geom_jitter() +
  labs(
    title = "Monthly Changes of Variables",
    x = "Months of August & September",
    y = "Variable's Values") +
  facet_wrap(vars(category), scales = "free_y")
  theme_bw()
  
# It is clear that FFMC, DC, DMC, wind, and temp have positive relationships with the time variable "month"
  
# To better see such patterns, let's limit the "x" xis to the the months of aug & sep

forest.df_long %>%
  ggplot(aes(x = month, y = value)) +
  geom_point() +
  geom_jitter() +
  labs(
    title = "Monthly Changes of Variables",
    x = "Months of August & September",
    y = "Variable's Values") +
  xlim("aug", "sep") +
  facet_wrap(vars(category), scales = "free_y")
  theme_bw()
```

# look at the relationship between these variables (FFMC, DMC, DC, ISI, temp, RH, wind, and rain) and area

```{r}
forest.df_long %>%
  ggplot(aes(x = value, y = area)) +
  geom_point() +
  geom_jitter() +
  labs(
    title = "Relationships between the Variables & Area",
    x = "Variable's Values",
    y = "Area Burned in hactare") +
  facet_wrap(vars(category), scales = "free_x")
  theme_bw()
```

## To better get insights into the relationships, it is recommended to address the outliers in the "area" variable

```{r}

forest.df_long %>%
  filter(area < 300 & area != 0) %>%
  ggplot(aes(x = value, y = area)) +
  geom_point( ) +
  geom_jitter() +
  labs(
    title = "Relationships between the Variables & Area",
    x = "Variable's Values",
    y = "Area Burned in hactare") +
  facet_wrap(vars(category), scales = "free_x")
  theme_bw()

```



