---
title: 'Guided Project: Analyzing Movie Ratings'
author: "Yassir"
date: "2023-07-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

## In this guided project, I will perform web data extraction from the IMDB which is the world's most popular and authoritative source for movie, TV, and celebrity content.
## The goal is to gather information from the 30 most popular movies released between March and July 2020. I intend to extract diverse details such as the movie's title, description, actors, director, genre, runtime, and ratings. I further examine if there is a correlation between ratings and user votes.

# Loading the Web Page
```{r}
# loading rvest, dplyr, and ggplot2 packages
library(rvest)
library(tidyverse)

# specifying the url
wp_content <- read_html("http://dataquestio.github.io/web-scraping-pages/IMDb-DQgp.html")
```

# String Manipulation Reminder
```{r}
# Converting "10.50" into numeric
as.numeric("10.50")

# Converting the vector c("14.59", "3.14", "55") into numeric
as.numeric(c("14.59", "3.14", "55"))

# Parsing the vector c("14 min", "17,35", "(2012)", "1,2,3,4") into numeric
parse_number(c("14 min", "17,35", "(2012)", "1,2,3,4"))

# Removing whitespaces at the beginning and end of " Spaces before and after should disappear    "
str_trim(" Spaces before and after should disappear    ")

```
# Extracting Elements from the Header
```{r}
# Extract the movies' titles
# CSS selector is identified using the SelectorGadget
movies_titles <- wp_content %>%
  html_nodes(".lister-item-header a") %>% 
  html_text()
print(movies_titles)

# Extract the movies' years.
movies_years <- wp_content %>%
  html_nodes(".text-muted.unbold") %>%
  html_text()
print(movies_years)
# eliminating the parentheses using the parse_number
movies_years <- parse_number(movies_years)

# convert the values from character to numeric data type
movies_years <- as.numeric(movies_years)
print(movies_years)

```
# Extracting the Movies' Features
```{r}
# extracting the movies_runtimes
movies_runtimes <- wp_content %>%
  html_nodes(".runtime") %>%
  html_text()
print(movies_runtimes) 

# parsing the numbers
movies_runtimes <- parse_number(movies_runtimes)

# converting the values from characters into numberic data type
movies_runtimes <- as.numeric(movies_runtimes)
print(movies_runtimes)

# Extract the movies' genres
movies_genres <- wp_content %>%
  html_nodes(".genre") %>%
  html_text()
print(movies_genres) 

# Removing white spaces at the end of genre characters
movies_genres <- str_trim(movies_genres)
print(movies_genres)

```

# Extracting the Movies' Ratings
```{r}
# Extracting the movies' user ratings.

movies_user_ratings <- wp_content %>%
  html_nodes(".ratings-imdb-rating") %>%
  html_attr("data-value")
print(movies_user_ratings) 

# Converting the user ratings from character to numeric data type
movies_user_ratings <- as.numeric(movies_user_ratings)
print(movies_user_ratings)

# Extracting the movies' metascores
movies_metascores <- wp_content %>%
  html_nodes(".metascore") %>%
  html_text( )
print(movies_metascores) 

# Removing whitespaces at the end of metascores 
movies_metascores <- str_trim(movies_metascores)

# converting them into numeric data type
movies_metascores <- as.numeric(movies_metascores)
print(movies_metascores)
```
# Extracting the Movies' Votes
```{r}
# Extracting the movies' votes
movies_votes <- wp_content %>%
  html_nodes(".sort-num_votes-visible :nth-child(2)") %>%
  html_text( )
print(movies_votes) 

# parsing the numbers
movies_votes <- parse_number(movies_votes)
print(movies_votes)

```

# Dealing with Missing Values
```{r}
# Copying & pasting the created append_vector()
append_vector <- function(vector, inserted_indices, values){

  ## Creating the current indices of the vector
  vector_current_indices <- 1:length(vector)

  ## Adding small amount of values (between 0 and 0.9) to the `inserted_indices`
  new_inserted_indices <- inserted_indices + seq(0, 0.9, length.out = length(inserted_indices))

  ## Appending the `new_inserted_indices` to the current vector indices
  indices <- c(vector_current_indices, new_inserted_indices)

  ## Ordering the indices
  ordered_indices <- order(indices)

  ## Appending the new value to the existing vector
  new_vector <- c(vector, values)

  ## Ordering the new vector wrt the ordered indices
  new_vector[ordered_indices]
}

# Using the append_vector() function to insert NA into the metascores vector after the positions 1, 1, 1, 13, and 24

movies_metascores <- append_vector(movies_metascores,c(1,1,1,13,24), NA)
print(movies_metascores)

# Remove the 17th element from the vectors: titles, years, runtimes, genres, and metascores

movies_genres <- movies_genres[-17]
movies_metascores <- movies_metascores[-17]
movies_runtimes <- movies_runtimes[-17]
movies_titles <- movies_titles[-17]
movies_years <- movies_years[-17]
```

# Putting It All Together and Visualizing
```{r}
# Creating a dataframe with the data we previously extracted: titles, years, runtimes, genres, user ratings, metascores, and votes.
  # Keep only the integer part of the user ratings using the floor() function. For example, 3.4 becomes 3

library(tibble)
movies_df <- tibble("titles" = movies_titles,
                    "years" = movies_years,
                    "runtimes" = movies_runtimes,
                    "genres" = movies_genres,
                    "user_ratings" = floor(movies_user_ratings),
                    "metascores" = movies_metascores,
                    "votes" = movies_votes)

print(movies_df)

# Creating a boxplot that shows the number of votes against the user rating.
ggplot(data = movies_df,
       aes(x = user_ratings, y = votes, group = user_ratings)) +
  geom_boxplot()

```

