---
title: 'Guided Project: Designing and Creating a Database'
author: "Yassir"
date: "2023-04-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Exploring the datasets

- loading tidyverse package
```{r}
library(tidyverse)
```
- importing the datasets
```{r}
game_log <- read_csv("game_log.csv")
park_codes <- read_csv("park_codes.csv")
person_codes <- read_csv("person_codes.csv")
team_codes <- read_csv("team_codes.csv")
```
1. explore the datasets and write a brief description of each file, and how they intersect with the game_log file?
```{r}
dim(game_log) # 161 columns with 171907 rows
head(game_log)
```
- The game_log file consists of 161 columns with 171907 rows
- The table has no primary key
- The table contains information on games, teams, players, umpires,and games' outcomes

```{r}
dim(park_codes) # 9 columns with 252 rows
head(park_codes)
```
- The table contains 9 columns and 252 rows
- The table has a primary key *park_id* which could act as a foreign key in the log_game table
- The table contains information on parks, cities, states, start and end dates, leagues, and notes.

```{r}
dim(person_codes) # 7 columns with 20494 rows
head(person_codes)
```
- The table has 7 columns with 20494 rows
- The table has a primary key
- The table contains first and last names 
- The table has information on the debut dates of players, managers, coaches, and umpires

```{r}
dim(team_codes) # 8 columns with 150 rows
head(team_codes)
```
- The table has 8 columns with 150 rows
- The table has a primary key
- The table contains information on league, start and end date, nickname, franch id. The league column is connected to league of the *game_log's* column

2. What each defensive position number represents in the *game_log_fields.txt* ?
- I used chatgpt to answer this question:
  -  Pitcher
  - catcher
  - first baseman
  - second baseman
  - third baseman
  - shortstop
  - left field
  - center field
  - right field
  - 10th position is not clear yet!
  
2. Write a short paragraph on the values in the various league fields, and which leagues they represent ?

- Using chatgpt, the values refer to the following leagues
  - NL: National League
  - AA: American Association
  - UA: Union Association
  - PL: Players' League 
  - AL: American League
  - FL: Federal League

## Importing Data into SQLite

1. Use dbWriteTable() to create tables for each of our csv files into a new SQLite database, mlb.db
  
```{r}
library(RSQLite)
conn <- dbConnect(SQLite(), "mlb.db")

dbWriteTable(conn = conn, name = "game_log",
             value = game_log, 
             row.names = FALSE, header = TRUE)

dbWriteTable(conn = conn, name = "park_codes",
             value = park_codes,
             row.names = FALSE, header = TRUE)

dbWriteTable(conn = conn, name = "person_codes",
             value = person_codes,
             row.names = FALSE, header = TRUE)

dbWriteTable(conn = conn, name = "team_codes",
             value = team_codes,
             row.names = FALSE, header = TRUE)

dbListTables(conn) # all tables are in there
```

2. Create a new column in the game_log table called game_id:

  - Use SQL string concatenation to update the new columns with a unique ID using the Retrosheet format.
  - Once you write the query, you can use dbExecute() to perform the query

```{r}
# Add new column *game_id* to game_log table
alter_game_log <- "ALTER TABLE game_log
                    ADD COLUMN game_id TEXT;"
# Use dbExecute () to perform the query

conn <- dbConnect(SQLite(), "mlb.db")

dbExecute(conn, alter_game_log)

# Use SQL string concatenation to update the new column and execute the querry
update_game_log <- "
 UPDATE game_log
    SET game_id = date || h_name ||         number_of_game
  WHERE game_id IS NULL;"

dbExecute(conn, update_game_log)

# Check if the game_id is added
head(dbGetQuery(conn, "select game_id, date,
                   h_name, number_of_game
                   from game_log;"))

```

## Looking for Normalization Opportunities

1. Looking at the various files, look for opportunities to normalize the data 

  - Repetition of the same type of information in columns (106-132) in the game_log
  - Repetition of the same type of information in columns (133-159) in the game log
  - To remove such repetitions, we could stack these columns on top of each other and creating new columns
  - Players' names could be removed from game log because the person_codes has a list of names and IDs
  - In team_codes, the start, end and sequence columns could be obtained from game log data
  - In person_codes, all the debut dates could be extracted be using game log data.
  - There are lots of places in game log where we have a player ID followed by the players name. We will be able to remove this and use the name data in person_codes
  - The debut coaches column should be removed because the game_log has no data on coaches
  - The start, end, and league columns contain data that is found in the main game log and can be removed.

## Creating Tables Without Foreign Key Relations

```{r}
# Create the person table with columns and primary key as shown in the schema diagram.

create_person <- "
CREATE TABLE IF NOT EXISTS person (
      person_id TEXT PRIMARY KEY,
      first_name TEXT,
      last_name TEXT
  );"
dbExecute(conn, create_person)

# Insert the data from the person_codes table.

insert_to_person <- "INSERT OR IGNORE INTO person
SELECT id,
     last,
     first
  from person_codes;"
dbExecute(conn, insert_to_person)

# Write a query to display the first few rows of the table.
head(dbGetQuery(conn, "select * from person
                limit 3;"))

```
```{r}
# Create the park table with columns and primary key as shown in the schema diagram.

create_park <- "
CREATE TABLE IF NOT EXISTS park (
      park_id TEXT PRIMARY KEY,
      name TEXT,
      nickname TEXT,
      city TEXT,
      state TEXT,
      notes TEXT
  );"
dbExecute(conn, create_park)

# Insert the data from the park_codes table.

insert_to_park <- "INSERT OR IGNORE INTO park
SELECT park_id,
     name,
     aka,
     city,
     state,
     notes
  from park_codes;"
dbExecute(conn, insert_to_park)

# Write a query to display the first few rows of the table.
head(dbGetQuery(conn, "select * from park
                limit 3;"))

```
```{r}
# create a table to store league names.
create_league <- '
CREATE TABLE IF NOT EXISTS league(
league_id TEXT PRIMARY KEY,
name TEXT);'
dbExecute(conn, create_league)

# inserting the data
insert_to_league <- '
  INSERT OR IGNORE INTO league
  VALUES
      ("NL", "National League"),
      ("AL", "American League"),
      ("AA", "American Association"),
      ("FL", "Federal League"),
      ("PL", "Players League"),
      ("UA", "Union Association");'

dbExecute(conn, insert_to_league)

# Write a query to display the first few rows of the table.
head(dbGetQuery(conn, "select * from league
                limit 3;"))

```
```{r}

# Import the data from appearance_type.csv
appearance_drop <- "DROP TABLE IF EXISTS appearance_type;"
dbExecute(conn, appearance_drop)
appearance_type <- read_csv("appearance_type.csv")

appearance_type  <-  read_csv('appearance_type.csv')
dbWriteTable(conn = conn, name = "appearance_type", 
             value = appearance_type,
             row.names = FALSE, header = TRUE)

# Write a query to display the table
head(dbGetQuery(conn, "select * from appearance_type
limit 3;"))


```
## Adding the Team and Game Tables

```{r}
# Create the team table with columns, primary key, and foreign key as shown in the schema diagram.
create_team <- '
CREATE TABLE IF NOT EXISTS team(
team_id TEXT PRIMARY KEY,
league_id TEXT,
city TEXT,
nickname TEXT,
franch_id,
FOREIGN KEY (league_id) REFERENCES league(league_id));'
dbExecute(conn, create_team)

# Insert the data from the team_codes table
insert_to_team <- 'INSERT OR IGNORE INTO team
SELECT team_id,
     league,
     city,
     nickname,
     franch_id
  from team_codes;'
dbExecute(conn, insert_to_team)
# Write a query to display the first few rows of the table
head(dbGetQuery(conn, 'select * from team
                       limit 3;'))
```


```{r}
# Create the game table with columns, primary key, and foreign key as shown in the schema diagram
create_game <- '
CREATE TABLE IF NOT EXISTS game(
game_id TEXT PRIMARY KEY,
date TEXT,
number_of_game INTEGER,
park_id TEXT,
length_outs INTEGER,
day BOOLEAN,
completion TEXT,
fortfeit TEXT,
protest TEXT,
attendance INTEGER,
lenght_minutes INTEGER,
additional_info TEXT,
acquisition_info TEXT,
FOREIGN KEY (park_id) REFERENCES park(park_id));'
dbExecute(conn, create_game)

# Insert the data from the game_log table
insert_to_game <- 'INSERT OR IGNORE INTO game
SELECT 
      game_id,
      date,
      number_of_game,
      park_id,
      length_outs,
      CASE
          WHEN day_night = "D" THEN 1
          WHEN day_night = "N" THEN 0
          ELSE NULL
          END
          AS day,
      completion,
      forfeit,
      protest,
      attendance,
      length_minutes,
      additional_info,
      acquisition_info
  FROM game_log;'
dbExecute(conn, insert_to_game)

# Write a query to display the first few rows of the table.
head(dbGetQuery(conn, 'select * from game
                       limit 3;'))
```

## Adding the Team Appearance Table

```{r}
# Create the team_appearance table with columns, primary key, and foreign keys as shown in the schema diagram.
create_team_appearance <- "
CREATE TABLE IF NOT EXISTS team_appearance (
      team_id TEXT,
      game_id TEXT,
      home BOOLEAN,
      league_id TEXT,
      score INTEGER,
      line_score TEXT,
      at_bats INTEGER,
      hits INTEGER,
      doubles INTEGER,
      triples INTEGER,
      homeruns INTEGER,
      rbi INTEGER,
      sacrifice_hits INTEGER,
      sacrifice_flies INTEGER,
      hit_by_pitch INTEGER,
      walks INTEGER,
      intentional_walks INTEGER,
      strikeouts INTEGER,
      stolen_bases INTEGER,
      caught_stealing INTEGER,
      grounded_into_double INTEGER,
      first_catcher_interference INTEGER,
      left_on_base INTEGER,
      pitchers_used INTEGER,
      individual_earned_runs INTEGER,
      team_earned_runs INTEGER,
      wild_pitches INTEGER,
      balks INTEGER,
      putouts INTEGER,
      assists INTEGER,
      errors INTEGER,
      passed_balls INTEGER,
      double_plays INTEGER,
      triple_plays INTEGER,
      PRIMARY KEY (team_id, game_id),
      FOREIGN KEY (team_id) REFERENCES team(team_id),
      FOREIGN KEY (game_id) REFERENCES game(game_id),
      FOREIGN KEY (team_id) REFERENCES team(team_id)
  );
"
dbExecute(conn, create_team_appearance)

# Insert the data from the game_log table, using a UNION clause to combine the data from the column sets for the home and away teams.
insert_to_team_appearance <- "
  INSERT OR IGNORE INTO team_appearance
      SELECT
          h_name,
          game_id,
          1 AS home,
          h_league,
          h_score,
          h_line_score,
          h_at_bats,
          h_hits,
          h_doubles,
          h_triples,
          h_homeruns,
          h_rbi,
          h_sacrifice_hits,
          h_sacrifice_flies,
          h_hit_by_pitch,
          h_walks,
          h_intentional_walks,
          h_strikeouts,
          h_stolen_bases,
          h_caught_stealing,
          h_grounded_into_double,
          h_first_catcher_interference,
          h_left_on_base,
          h_pitchers_used,
          h_individual_earned_runs,
          h_team_earned_runs,
          h_wild_pitches,
          h_balks,
          h_putouts,
          h_assists,
          h_errors,
          h_passed_balls,
          h_double_plays,
          h_triple_plays
      FROM game_log
  
  UNION
  
      SELECT    
          v_name,
          game_id,
          0 AS home,
          v_league,
          v_score,
          v_line_score,
          v_at_bats,
          v_hits,
          v_doubles,
          v_triples,
          v_homeruns,
          v_rbi,
          v_sacrifice_hits,
          v_sacrifice_flies,
          v_hit_by_pitch,
          v_walks,
          v_intentional_walks,
          v_strikeouts,
          v_stolen_bases,
          v_caught_stealing,
          v_grounded_into_double,
          v_first_catcher_interference,
          v_left_on_base,
          v_pitchers_used,
          v_individual_earned_runs,
          v_team_earned_runs,
          v_wild_pitches,
          v_balks,
          v_putouts,
          v_assists,
          v_errors,
          v_passed_balls,
          v_double_plays,
          v_triple_plays
      from game_log;
"
dbExecute(conn, insert_to_team_appearance)

# Write a query to verify that your data was inserted correctly.
head(dbGetQuery(conn, 'SELECT * FROM team_appearance
  WHERE game_id = (
                   SELECT MIN(game_id) from game
                  )
     OR game_id = (
                   SELECT MAX(game_id) from game
                  )
  ORDER By game_id, home'))

```

## Adding the Person Appearance Table

```{r}
# Create the person_appearance table with columns, primary key, and foreign keys as shown in the schema diagram.

drop_person_appearance <- "
DROP TABLE IF EXISTS person_appearance"
dbExecute(conn, drop_person_appearance)

create_person_appearance <- "
CREATE TABLE person_appearance (
      appearance_id INTEGER PRIMARY KEY,
      person_id TEXT,
      team_id TEXT,
      game_id TEXT,
      appearance_type_id TEXT,
      FOREIGN KEY (person_id) REFERENCES person(person_id),
      FOREIGN KEY (team_id) REFERENCES team(team_id),
      FOREIGN KEY (game_id) REFERENCES game(game_id),
      FOREIGN KEY (appearance_type_id) REFERENCES appearance_type(appearance_type_id)
  );
"
dbExecute(conn, create_person_appearance)

# Insert the data from the game_log table, using UNION clauses to combine the data from the columns for managers, umpires, pitchers, and awards.
insert_to_person_appearance <- '
  INSERT OR IGNORE INTO person_appearance (
      game_id,
      team_id,
      person_id,
      appearance_type_id
  ) 
      SELECT
          game_id,
          NULL,
          hp_umpire_id,
          "UHP"
      FROM game_log
      WHERE hp_umpire_id IS NOT NULL    
  
  UNION
  
      SELECT
          game_id,
          NULL,
          [1b_umpire_id],
          "U1B"
      FROM game_log
      WHERE "1b_umpire_id" IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          NULL,
          [2b_umpire_id],
          "U2B"
      FROM game_log
      WHERE [2b_umpire_id] IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          NULL,
          [3b_umpire_id],
          "U3B"
      FROM game_log
      WHERE [3b_umpire_id] IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          NULL,
          lf_umpire_id,
          "ULF"
      FROM game_log
      WHERE lf_umpire_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          NULL,
          rf_umpire_id,
          "URF"
      FROM game_log
      WHERE rf_umpire_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          v_name,
          v_manager_id,
          "MM"
      FROM game_log
      WHERE v_manager_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          h_name,
          h_manager_id,
          "MM"
      FROM game_log
      WHERE h_manager_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          CASE
              WHEN h_score > v_score THEN h_name
              ELSE v_name
              END,
          winning_pitcher_id,
          "AWP"
      FROM game_log
      WHERE winning_pitcher_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          CASE
              WHEN h_score < v_score THEN h_name
              ELSE v_name
              END,
          losing_pitcher_id,
          "ALP"
      FROM game_log
      WHERE losing_pitcher_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          CASE
              WHEN h_score > v_score THEN h_name
              ELSE v_name
              END,
          saving_pitcher_id,
          "ASP"
      FROM game_log
      WHERE saving_pitcher_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          CASE
              WHEN h_score > v_score THEN h_name
              ELSE v_name
              END,
          winning_rbi_batter_id,
          "AWB"
      FROM game_log
      WHERE winning_rbi_batter_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          v_name,
          v_starting_pitcher_id,
          "PSP"
      FROM game_log
      WHERE v_starting_pitcher_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          h_name,
          h_starting_pitcher_id,
          "PSP"
      FROM game_log
      WHERE h_starting_pitcher_id IS NOT NULL;
'
dbExecute(conn, insert_to_person_appearance)

# Use a loop with string formatting to insert the data for offensive and defensive positions from the game_log table
for (letter in c("h", "v")) {
  for (num in 1:9) {
    template <- '
      INSERT INTO person_appearance (
          game_id,
          team_id,
          person_id,
          appearance_type_id
      ) 
          SELECT
              game_id,
              %s_name,
              %s_player_%f_id,
              "O%f"
          FROM game_log
          WHERE %s_player_%f_id IS NOT NULL
      
      UNION
      
          SELECT
              game_id,
              %s_name,
              %s_player_%f_id,
              "D" || CAST(%s_player_%f_def_pos AS INT)
          FROM game_log
          WHERE %s_player_%f_id IS NOT NULL;
    '
    # replace all of the %s and %f with the correct letter number
    template <- gsub("%s", letter, template, fixed = TRUE)
    template <- gsub("%f", num, template, fixed = TRUE)
    
    dbExecute(conn, template)
  }
}

# Write a query to verify that your data was inserted correctly.
head(dbGetQuery(conn, 'SELECT * FROM person_appearance
                       limit 3;'))
```

## Removing the Original Tables

```{r}
# Drop the tables we created to hold our unnormalized data:
  # game_log.
  # park_codes.
  # team_codes.
  # person_codes.
# first we check the current tables
dbListTables(conn)
# let's drop the tables
dbExecute(conn, 'DROP TABLE game_log')
dbExecute(conn, 'DROP TABLE park_codes')
dbExecute(conn, 'DROP TABLE team_codes')
dbExecute(conn, 'DROP TABLE person_codes')

# let's make sure these tables are dropped
dbListTables(conn)

# Finally, let's disconnect from out database
dbDisconnect(conn)
```