---
title: 'Kaggle : Kaggle Employee dataset with Multivariate Logistic Regression'
author: "Yassir"
date: "2023-10-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This dataset contains information about employees in a company, including their educational backgrounds, work history, demographics, and employment-related factors

# Exploring the dataset
```{r}
#loading tidyverse
# importing the dataset
library(tidyverse)
employee_df <- read_csv("Employee.csv")

dim(employee_df) # 9 variables & 4653 rows
head(employee_df)
str(employee_df)
glimpse(employee_df)
```
# let's check if there is missing values
```{r}
# no missing values
sum(is.na(employee_df))
```
# let's transform some variables into factors to better visualize them
```{r}
# let's use as_factor from forcats and mutate from dplyr
employee_df <- employee_df %>% mutate(LeaveOrNot = as_factor(LeaveOrNot))

employee_df <- employee_df %>% mutate(Gender = as_factor(Gender))

employee_df <- employee_df %>%
               mutate(Education = as_factor(Education))

```

# What is the distribution of educational qualifications among employees?
```{r}
# let's visualize the distribution of educational qualifications among employees
ggplot(data = employee_df, aes(x = Education, fill = Education)) +
  geom_bar()

# let's visualize the educational qualifications using percentages
ggplot(data = employee_df, 
       aes(x = Education, y = ..prop.. * 100, group = 1)) +
geom_bar() +
  labs(title = "Distribution of Educational Qualifications Among Employees",
       x = " Employees' Educational Qualification",
       y = "Percentages")

# Based on the bar chart, almost 80% of employees have bachelor degrees while around 20% and 5% of employees have master and PhD degrees respectively.

# we can solve it using mutate from dplyr

employee_df %>%
  count(Education) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  ggplot(aes(x = Education, y = percentage, fill = Education)) +
  geom_bar(stat = "identity") +
  labs(title = "Distribution of Educational Qualifications Among Employees",
       x = "Employees' Educational Qualification",
       y = "Percentages") +
  scale_y_continuous(labels = scales::percent_format())

```

# How does the length of service (Joining Year) vary across different cities?
```{r}
# Create a grouped histogram
ggplot(data = employee_df, aes(x = JoiningYear, fill = City)) +
  geom_histogram(binwidth = 0.5,position = "dodge") +
  labs(title = "Length of Service Across Different Cities",
       x = "Joining Year",
       y = "Frequency") 
# Employees in Bangalore city joined longer compared to Pune and New Delhi respectively.

# we can also use box plot to visualize the length of service across the different citites
ggplot(employee_df, aes(x = City, y = JoiningYear, fill = City)) +
  geom_boxplot() +
  labs(title = "Length of Service Across Different Cities",
       x = "City",
       y = "Joining Year")
```

# Is there a correlation between Payment Tier and Experience in Current Domain?
```{r}
# We apply the spearman method to find the correlations between ordinal variables 

# let's rank the ExperienceInCurrentDomain variable
experience_rank <- rank(employee_df$ExperienceInCurrentDomain)

# let's find the Spearman correlation
cor(employee_df$PaymentTier, experience_rank, method = "spearman")

#  0.01519145
# Based on the above correlation coefficient, it suggests that there is very little evidence of a monotonic relationship between the two variables
```

# What is the gender distribution within the workforce?
```{r}
unique(employee_df$Gender)
ggplot(data = employee_df, aes(x = (Gender), fill = Gender)) +
  geom_bar() +
  labs(title = "Distribution of Employees by Gender") +
  scale_fill_manual(values = c("blue", "pink"), name = "Gender", labels = c("Male", "Female"))

```

# Are there any patterns in leave-taking behavior among employees?
```{r}
# patterns based on Education
# employees with master degrees are more likely to leave the firm

ggplot(data = employee_df, aes(x = employee_df$Education, fill =  employee_df$LeaveOrNot)) +
  geom_bar(position ="dodge") +
  labs(title = "Leave-Taking Behavior Among Employees by Education Level",
       x = "Education Level",
       y = "Count",
       fill = "Leave or Not") +
  scale_fill_discrete(name = "Leave or Not", labels = c("Stay", "Leave"))
```

```{r}
# patterns based on gender
# females employees tend to leave the firm more than males do

ggplot(data = employee_df, aes(x = employee_df$Gender, fill =  employee_df$LeaveOrNot)) +
  geom_bar(position ="dodge") +
  labs(title = "Leave-Taking Behavior Among Employees by Gender",
       x = "Gender",
       y = "Count",
       fill = "Leave or Not") +
  scale_fill_discrete(name = "Leave or Not", labels = c("Stay", "Leave"))

```

```{r}
# patterns based on payement tier

# employees with payment tier 2 tends to leave the firm more than those who stay with payment tier 2

ggplot(data = employee_df, aes(x = employee_df$PaymentTier, fill =  employee_df$LeaveOrNot)) +
  geom_bar(position ="dodge") +
  labs(title = "Leave-Taking Behavior Among Employees by Payment Tier",
       x = "Payment Tier",
       y = "Count",
       fill = "Leave or Not") +
  scale_fill_discrete(name = "Leave or Not", labels = c("Stay", "Leave")) 

```

```{r}
# patterns based on age
# Employees whose age range between 24-29 are more likely to leave the firm

ggplot(data = employee_df, aes(x = employee_df$Age, fill =  employee_df$LeaveOrNot)) +
  geom_histogram(position ="dodge", bins = 30) +
  labs(title = "Leave-Taking Behavior Among Employees by Age",
       x = "Age",
       y = "Count",
       fill = "Leave or Not") +
  scale_fill_discrete(name = "Leave or Not", labels = c("Stay", "Leave")) 

# we can use box plot as well!

ggplot(employee_df, aes(x = LeaveOrNot, y = Age, fill = LeaveOrNot)) +
  geom_boxplot() +
  labs(title = "Leave-Taking Behavior by Age",
       x = "Leave or Not",
       y = "Age") +
  scale_fill_discrete(name = "Leave or Not", labels = c("Stay", "Leave"))

```

```{r}
# patterns based on City
# employees in Pune city tend to leave more than those employees who stay in Pune city

ggplot(data = employee_df, aes(x = City, fill =  employee_df$LeaveOrNot)) +
  geom_bar(position = "dodge") +
  labs(title = "Leave-Taking Behavior Among Employees by City",
       x = "City",
       y = "Count",
       fill = "Leave or Not") +
  scale_fill_discrete(name = "Leave or Not", labels = c("Stay", "Leave")) 


```

```{r}
# patterns based on joining year
# employees left the firm in 2018

ggplot(data = employee_df, aes(x = employee_df$JoiningYear, fill =  employee_df$LeaveOrNot)) +
  geom_bar(position ="dodge") +
  labs(title = "Leave-Taking Behavior Among Employees by Joining Year",
       x = "Joining Year",
       y = "Count",
       fill = "Leave or Not") +
  scale_fill_discrete(name = "Leave or Not", labels = c("Stay", "Leave")) 


```

let's do multivariate logistic regression model to predict the "LeaveOrNot" outcome variable
```{r}
# Fit logistic regression model
model <- glm(LeaveOrNot ~ Education + City + Gender + Age + PaymentTier, data = employee_df, family = "binomial")


# Summary of the logistic regression model
summary(model)

```